A Linear Solution of the Linearity Challenge

### 2025-12-25 paper/calculus.tex
global revision

The interpretation of linear logic propositions as protocols is fairly
usual~\cite{Wadler14}: the multiplicative constants $\Bot$ and $\One$ describe
protocols that terminate with the input/output of a signal; the additive
connectives $A \With B$ and $\A \Plus B$ describe protocols having two possible
continuations of type $A$ and $B$ which are selected/offered; the multiplicative
connectives $A \Par B$ and $A \Ten B$ describe bifurcating protocols with two
continuations of type $A$ and $B$. The additive constants $\Zero$ and $\Top$ do
not describe useful protocols, but they can play the role of smallest/largest
type in type systems that support a notion of subtyping~\citep{HornePadovani24}. 
%
The quantifiers $\forall X.A$ and $\exists X.A$ describe the input/output of a
type $X$ with a continuation $A$ and are useful to describe parametric protocol
polymorphism.
%
Finally, the ``of course'' modality $\OfCourse A$ and the ``why not'' modality
$\WhyNot A$ describe shared channels on which servers and clients accept and
request connections of type $A$.

@@

The syntax of processes makes use of an infinite set of \emph{channels}, ranged
over by $x$, $y$ and $z$, and is shown in~\Cref{tab:syntax}.
%
A link $\Link\x\y$ denotes the merging of the channels $x$ and $y$, so that
every message sent on one of the channels is forwarded to the other. As
discussed in the literature~\cite{LindleyMorris15}, this form is useful for
modeling higher-order sessions, that is the exchange of an existing channel on
another channel.
%
The processes $\Wait\x.P$ and $\Close\x$ respectively model the input and output
of a termination signal on the channel $x$. The latter process has no
continuation, while the former process continues as $P$ once the signal has been
received.
%
The process $\Fail\x$ can be used to denote a failure concerning the channel
$x$.
%
The processes $\Case\x\z{Q_1}{Q_2}$ and $\Select\x{\InTag_i}\z.P$ for $i=1,2$
describe the input/output of a label $\InTag_i$ along with a fresh continuation
$z$ on the channel $x$. The receiver continues as $Q_i$ depending on the
received label $\InTag_i$. In practice it is useful to consider arbitrary
(domain-specific) sets of labels instead of the canonical $\InTag_1$ and
$\InTag_2$. This generalisation has no substantial impact on the formalisation,
so we stick to binary choices in line with the usual presentations of linear
logic.
%
The processes $\Join\x\y\z.R$ and $\Fork\x\y\z{P}{Q}$ describe the input/output
of a pair of fresh channels $y$ and $z$ on the channel $x$. The former process
has a single continuation $R$ that can use both $y$ and $z$. The latter process
forks into $P$ and $Q$, each using $y$ and $z$ respectively. These forms are
often used in conjunction with links to model higher-order sessions.
%
The processes $\Forall\x\X\z.P$ and $\Exists\x{A}\z.P$ describe the input/output
of a type on the channel $x$ along with a fresh continuation $z$.

@@

\begin{example}
    \label{ex:booleans}
    We can represent the type of a boolean value being produced as the
    proposition $\Bool \eqdef \One \Plus \One$ and that of a boolean value being
    consumed as its dual $\dual\Bool = \Bot \With \Bot$. Following these types,
    the boolean constants can be modeled by the processes
    \[
        \TrueP(x) \eqdef \Select\x\LeftTag{x'}.\Close{x'}
        \qquad
        \FalseP(x) \eqdef \Select\x\RightTag{x'}.\Close{x'}
    \]
    and the boolean negation function by the process
    \[
        \NotP(x,y) \eqdef
        \BigCase\x{x'}{
            \Wait{x'}.\FalseP(y)
        }{
            \Wait{x'}.\TrueP(y)
        }
    \]

    As an example, the composition $\Cut[\Bool]\x{\TrueP(x)}{\NotP(x,y)}$
    produces false on $y$.
    %
    Note the use of explicit continuations in these processes and the fact that
    each channel is used exactly once. The same processes in \CP would be
    written as
    \[
        \TrueP(x) \eqdef \Select\x\LeftTag{}.\Close\x
        \quad
        \FalseP(x) \eqdef \Select\x\RightTag{}.\Close\x
        \quad
        \NotP(x,y) \eqdef
        \BigCase\x{}{
            \Wait\x.\FalseP(y)
        }{
            \Wait\x.\TrueP(y)
        }
    \]
    where each channel is used multiple times to indicate the various
    input/output actions pertaining to the same session.
    %
    In general, the \CP version of an \LCP process can be obtained by reusing
    the same channel $x$ in place of the continuation $z$ in \Cref{tab:syntax}.
    %
    \eoe
\end{example}

@@

Structural pre-congruence is the least pre-congruence defined by the
\rulename{s-*} rules.
%
\SLink and \SComm assert that links and parallel compositions are commutative.
The remaining rules, when read from left to right, push a cut on $x$ underneath
the topmost prefix on $y$ of one of its sub-processes when $x \ne y$. These
rules are key to float input/output actions to the top-level of a process, so
that they can interact with corresponding complementary actions in the context
surrounding the process.
%
All these rules have implicit side-conditions (not shown in
\Cref{tab:semantics}) aimed at preserving the meaning of channels when binders
are moved around: terms entering or exiting the scope of a binder for $x$ must
not have free occurrences of $x$. This holds also for the type variable $X$ in
the rule \SForall.
%
We content ourselves with such informal description of these side conditions
given that we are going to formalise \Calculus later on.
%
Note also that there are two versions of \SForkL and \SForkR depending on which
of the two continuations (either $P$ or $Q$) contains free occurrences of the
restricted channel $x$.

There is no pre-congruence rule asserting that parallel composition is
associative. While this is a sound property of parallel composition, it turns
out to be unnecessary for proving the properties of \LCP that we consider. We
leave it out in light of the forthcoming formalisation of the calculus, in which
every simplification is welcome, although standard presentations of classical
linear logic admit such law~\cite{Wadler14}.
%
Another rule that deserves attention is \SServer. In this case, a server prefix
$\Server\y\u$ can be pulled out of a cut only provided that the other process in
the cut is also a server on the channel restricted by the cut.

Reduction is defined by the \rulename{r-*} rules.
%
The reduction rules \RLink, \RClose, \RSelect, \RFork, \RExists, \RConnect,
\RWeaken and \RContract are in one-to-one correspondence with the well-known
reductions of session calculi and the principal cut reductions of linear logic.
\RLink eliminates a link $\Link\x\y$ by substituting $y$ for $x$. \RClose closes
a session $x$ when a termination signal is exchanged on $x$. \RSelect and \RFork
respectively model the communication of a tag $\InTag_i$ and of a fresh channel
$y$. \RExists models the instantiation of a polymorphic variable $X$ as the
communication of a type $A$. We write $Q\subst{A}\X$ for the process obtained by
replacing every free occurrence of the type variable $X$ with $A$. Each of these
communications consumes the channel $x$ and possibly creates and transfers a
\emph{fresh continuation} $z$ on which the rest of the conversation takes place,
in line with the encoding of sessions into the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
%
The rules \RConnect, \RWeaken and \RContract describe interactions on shared
channels and respectively represent the connection between a server and a
client, the disposal of an unused server, and the duplication of a server. In
\RWeaken and \RContract we use some convenient (albeit slightly informal)
notation for denoting sequences of (pairwise distinct) channels and prefixes. In
particular, $\seqof\z$ stands for a sequence $z_1, \dots, z_n$ of channels,
$\Weaken{\seqof\z}.Q$ stands for $\Weaken{z_1}\dots\Weaken{z_n}.Q$ and
$\Contract{\seqof\z}{\seqof\z'}{\seqof\z''}.R$ stands for
$\Contract{z_1}{z_1'}{z_1''}\dots\Contract{z_n}{z_n'}{z_n''}.R$.

@@

Typing judgments have the form $\wtp{P}\Context$ meaning that the process $P$ is
well typed in the context $\Context$. Equivalently, the judgment indicates that
the sequent $\wtp{}\Context$ is derivable and $P$ is a \emph{proof term}
corresponding to the derivation for $\wtp{}\Context$.
%
The typing rules are shown in \Cref{tab:typing-rules}. Since they are in
one-to-one correspondence with -- and have exactly the same structure of -- the
proof rules of classical linear logic, the reader may refer to the standard
literature on linear logic or propositions as sessions~\citep{Wadler14} for
their interpretation.

@@

    The server accepts requests on the shared channel $x$. Each request
    established a new session, initially represented by the channel $y$. The
    session consists in receiving the type $X$ of the value being echoed
    followed by the value itself $u$. The value is then sent back unchanged and
    the session is terminated. Note the succession of different (linear)
    channels $y, y', y'', y'''$ that chain the session together.

@@

We say that $P$ is a \emph{thread} if it is anything but a cut. In other words,
a thread is either a link or a process that starts with an input/output action
of some sort. Note that a thread may contain cuts, but these cuts must all
guarded underneath the topmost action prefix of the thread.
%
We say that $P$ is \emph{observable} if $P \pcong Q$ for some thread $Q$. An
observable process is a process that exposes an action on a free channel and
therefore can interact through that channel, if put into some appropriate
context. We say that $P$ is \emph{reducible} if $P \red Q$ for some $Q$. A
reducible process may perform a reduction step.
%
We say that a process is \emph{alive} if it is either observable or reducible
and that it is a \emph{deadlock} if it is not alive.

>>>>

The interpretation of linear logic propositions as protocols is fairly
usual~\cite{Wadler14}: the multiplicative constants $\Bot$ and $\One$ describe
protocols that terminate with the input/output of a signal; the additive
connectives $A \With B$ and $\A \Plus B$ describe protocols having two possible
continuations of type $A$ and $B$ which are respectively offered/selected; the
multiplicative connectives $A \Par B$ and $A \Ten B$ describe bifurcating
protocols with two continuations of type $A$ and $B$ that are respectively
received and sent. The additive constants $\Zero$ and $\Top$ do not describe
useful protocols, but they can play the role of smallest/largest type in type
systems that support a notion of subtyping~\citep{HornePadovani24}. 
%
The quantifiers $\forall X.A$ and $\exists X.A$ describe the input/output of a
type $X$ with a continuation $A$ and are useful to describe parametric protocol
polymorphism.
%
Finally, the ``of course'' modality $\OfCourse A$ and the ``why not'' modality
$\WhyNot A$ describe shared channels on which servers and clients accept and
request connections of type $A$.

@@

The syntax of processes makes use of an infinite set of \emph{channels}, ranged
over by $x$, $y$ and $z$, and is shown in~\Cref{tab:syntax}.
%
A link $\Link\x\y$ denotes the merging of the channels $x$ and $y$, so that the
message sent on one of the channels is forwarded to the other. As discussed in
the literature~\cite{LindleyMorris15}, this form is useful for modeling
higher-order sessions, that is the exchange of an existing channel on another
channel.
%
The processes $\Wait\x.P$ and $\Close\x$ respectively model the input and output
of a termination signal on the channel $x$. The latter process has no
continuation, while the former process continues as $P$ once the signal has been
received.
%
The process $\Fail\x$ can be used to denote a failure concerning the channel
$x$.
%
The processes $\Case\x\z{Q_1}{Q_2}$ and $\Select\x{\InTag_i}\z.P$ for $i=1,2$
describe the input/output of a label $\InTag_i$ along with a fresh continuation
$z$ on the channel $x$. The receiver continues as $Q_i$ depending on the label
$\InTag_i$. In practice it is useful to consider arbitrary (domain-specific)
sets of labels instead of the canonical $\InTag_1$ and $\InTag_2$. This
generalisation has no substantial impact on the formalisation, so we stick to
binary choices in line with the usual presentations of linear logic.
%
The processes $\Join\x\y\z.R$ and $\Fork\x\y\z{P}{Q}$ describe the input/output
of a pair of fresh channels $y$ and $z$ on the channel $x$. The former process
has a single continuation $R$ that can use both $y$ and $z$. The latter process
forks into $P$ and $Q$, each using $y$ and $z$ respectively. These forms are
often used in conjunction with links to model higher-order sessions.
%
The processes $\Forall\x\X\z.P$ and $\Exists\x{A}\z.P$ describe the input/output
of a type on the channel $x$ along with a fresh continuation $z$.

@@

\begin{example}
    \label{ex:booleans}
    We can represent the type of a boolean value being produced as the protocol
    $\Bool \eqdef \One \Plus \One$ and that of a boolean value being consumed as
    its dual $\dual\Bool = \Bot \With \Bot$. Following these types, the boolean
    constants can be modeled by the processes
    \[
        \TrueP(x) \eqdef \Select\x\LeftTag{x'}.\Close{x'}
        \qquad
        \FalseP(x) \eqdef \Select\x\RightTag{x'}.\Close{x'}
    \]
    and the boolean negation function by the process
    \[
        \NotP(x,y) \eqdef
        \BigCase\x{x'}{
            \Wait{x'}.\FalseP(y)
        }{
            \Wait{x'}.\TrueP(y)
        }
    \]

    As an example, the composition $\Cut[\Bool]\x{\TrueP(x)}{\NotP(x,y)}$
    produces false on $y$.
    %
    Note the use of explicit continuations in these processes and the fact that
    each channel is used exactly once. The same processes in \CP would be
    written as
    \[
        \TrueP(x) \eqdef \Select\x\LeftTag{}.\Close\x
        \quad
        \FalseP(x) \eqdef \Select\x\RightTag{}.\Close\x
        \quad
        \NotP(x,y) \eqdef
        \BigCase\x{}{
            \Wait\x.\FalseP(y)
        }{
            \Wait\x.\TrueP(y)
        }
    \]
    where each channel is used multiple times to indicate the sequence of
    input/output actions pertaining to the same session.
    %
    In general, the \CP version of an \LCP process can be obtained by reusing
    the same channel $x$ in place of the continuation $z$ in \Cref{tab:syntax}.
    %
    \eoe
\end{example}

@@

Structural pre-congruence is the least pre-congruence defined by the
\rulename{s-*} rules.
%
\SLink and \SComm assert that links and parallel compositions are commutative.
The remaining rules, when read from left to right, push a cut on $x$ underneath
the topmost prefix on $y$ of one of its sub-processes when $x \ne y$. These
rules are key to float input/output actions to the top-level of a process, so
that they can interact with corresponding complementary actions in the
surrounding context.
%
All these rules have implicit side-conditions (not shown in
\Cref{tab:semantics}) aimed at preserving the meaning of channels when binders
are moved around: terms entering or exiting the scope of a binder for $x$ must
not have free occurrences of $x$. This holds also for the type variable $X$ in
the rule \SForall.
%
We content ourselves with such informal description of these side conditions
given that we are going to formalise \Calculus later on.
%
Note also that there are two versions of \SForkL and \SForkR depending on which
of the two continuations (either $P$ or $Q$) contains a free occurrence of the
restricted channel $x$.

There is no pre-congruence rule asserting that parallel composition is
associative. While this is a sound property of parallel composition, it turns
out to be unnecessary for proving the properties of \LCP that we consider. We
leave it out in light of the forthcoming formalisation of the calculus, in which
every simplification is welcome, although standard presentations of classical
linear logic admit such law~\cite{Wadler14,Doumane17}.
%
Another rule that deserves attention is \SServer. In this case, a server prefix
$\Server\y\u$ can be pulled out of a cut only provided that the other process in
the cut is also a server on the channel restricted by the cut.

Reduction is defined by the \rulename{r-*} rules.
%
The reduction rules \RLink, \RClose, \RSelect, \RFork, \RExists, \RConnect,
\RWeaken and \RContract are in one-to-one correspondence with the well-known
reductions of session calculi and the principal cut reductions of linear logic.
\RLink eliminates a link $\Link\x\y$ by substituting $y$ for $x$. \RClose closes
a session $x$ when a termination signal is exchanged on $x$. \RSelect and \RFork
respectively model the communication of a tag $\InTag_i$ and of a pair of fresh
continuation channels. \RExists models the instantiation of a polymorphic
variable $X$ as the communication of a type $A$. We write $Q\subst{A}\X$ for the
process obtained by replacing every free occurrence of the type variable $X$
with $A$. Each of these communications consumes the channel $x$ and possibly
creates and transfers \emph{fresh continuations} on which the rest of the
conversation takes place, in line with the encoding of sessions into the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
%
The rules \RConnect, \RWeaken and \RContract describe interactions on shared
channels and respectively represent the connection between a server and a
client, the disposal of an unused server, and the duplication of a server. In
\RWeaken and \RContract we use some convenient (albeit slightly informal)
notation for denoting sequences of (pairwise distinct) channels and prefixes. In
particular, $\seqof\z$ stands for a sequence $z_1, \dots, z_n$ of channels,
$\Weaken{\seqof\z}.Q$ stands for $\Weaken{z_1}\dots\Weaken{z_n}.Q$ and
$\Contract{\seqof\z}{\seqof\z'}{\seqof\z''}.R$ stands for
$\Contract{z_1}{z_1'}{z_1''}\dots\Contract{z_n}{z_n'}{z_n''}.R$.

@@

Typing judgments have the form $\wtp{P}\Context$ meaning that the process $P$ is
well typed in the context $\Context$. Equivalently, the judgment indicates that
the sequent $\wtp{}\Context$ is derivable and $P$ is a \emph{proof term}
corresponding to the derivation for $\wtp{}\Context$.
%
The typing rules are shown in \Cref{tab:typing-rules}. Since they are in
one-to-one correspondence with -- and have exactly the same structure of -- the
proof rules of classical linear logic, the reader may refer to the standard
literature on propositions as sessions~\citep{Wadler14} for their
interpretation.

@@

    The server accepts requests on the shared channel $x$. Each request creates
    a new session, initially represented by the channel $y$. The session
    consists in receiving the type $X$ of the value being echoed followed by the
    value itself $u$. The value is then sent back unchanged and the session is
    terminated. Note the succession of different (linear) channels $y, y', y'',
    y'''$ that chain the session together.

@@

We say that $P$ is a \emph{thread} if it is anything but a cut. In other words,
a thread is either a link or a process that starts with an input/output action
of some sort. Note that a thread may contain cuts, but these cuts must be
guarded underneath the topmost action prefix of the thread.
%
We say that $P$ is \emph{observable} if $P \pcong Q$ for some thread $Q$. An
observable process is a process that exposes an action on a free channel and
therefore can interact through that channel, if put into some appropriate
context. We say that $P$ is \emph{reducible} if $P \red Q$ for some $Q$. A
reducible process may perform a reduction step.
%
We say that a process is \emph{alive} if it is either observable or reducible
and that it is a \emph{deadlock} if it is not alive.

### 2025-12-25 paper/introduction.tex
revision
global revision

The Concurrent Calculi Formalisation Benchmark~\citep{CarboneEtAl24} is a
collection of challenges concerning the mechanisation of formal models of
concurrent and distributed programming languages.
%
These models often make use of distinctive features that set them apart from the
formal models of \emph{sequential} programming languages, such as the adoption
of substructural (linear, affine) type systems, the dynamic scope of first-class
channels in systems of communicating processes, the need for coinductive
definitions and proof methods for describing and reasoning on possibly infinite
behaviors.
%
The benchmark aims at identifying effective formalisation techniques that take
these features into account so as to foster the adoption of machine-checked
proofs in research work concerning concurrent and distributed programming
languages.

One of the challenges in the benchmark, henceforth called \emph{linearity
challenge}, concerns the formalisation of a \emph{minimal calculus of sessions}.
%
Sessions and session types~\citep{Honda93,HondaVasconcelosKubo98,HuttelEtAl16}
are established abstractions for the static analysis of distributed programs
based on peer-to-peer communication channels. Every session type system revolves
around three key ideas: (1) session endpoints are \emph{linear resources} that
cannot be discarded or duplicated without compromising some safety and liveness
properties of a program; (2) the type of a session endpoint is \emph{updated}
after each use to reflect the state of the interaction it describes; (3) peer
session endpoints are meant to be used in complementary ways so as to guarantee
the absence of communication errors and, to some extent, progress of the
interaction.
%
The linearity challenge is motivated by the observation that the proper
management of linear resources in a formalisation often requires a large number
of auxiliary definitions and technical results that divert from the main problem
under study~\cite{CarboneEtAl24}. One of the alleged culprits of such complexity
is \emph{context splitting}, namely the operation that partitions a typing
context in such a way that the linear resources described therein end up in only
one of the partitions. This observation has led to the exploration of various
alternative techniques including leftover typing~\cite{ZalakainDardha21}, the
use of linearity predicates~\cite{SanoKavanaghPientka23} and tagged
contexts~\cite{ZackonEtAl25}.

In this paper we approach the linearity challenge from a different angle:
\emph{instead of proposing new techniques that make it easy to formalise the
calculus in the challenge, we propose a (relatively) new calculus that is easy
to formalise with the existing techniques}.
%
More specifically, we describe a calculus called \LCP whose type system
coincides with the proof system of classical linear logic and that features
\emph{linear channels} instead of sessions. Linear channels are meant to be used
for a \emph{single} communication. Structured communications involving the
exchange of several messages within the same session is modeled in \LCP using
\emph{explicit continuations} according to the well-known encoding of (binary)
sessions into the linear
$\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
%
Overall, \LCP is a linear version of \CP~\cite{Wadler14} in which sessions are
encoded instead of being featured natively.

>>>>

The Concurrent Calculi Formalisation Benchmark~\citep{CarboneEtAl24} is a
collection of challenges concerning the mechanisation of formal models of
concurrent and distributed programming languages.
%
These models often make use of distinctive features that set them apart from the
formal models of sequential programming languages, such as the adoption of
substructural (linear, affine) type systems, the dynamic scope of first-class
channels in systems of communicating processes, the need for coinductive
definitions and proof methods for describing and reasoning on possibly infinite
behaviours.
%
The benchmark aims at identifying effective formalisation techniques that take
these features into account so as to foster the adoption of machine-checked
proofs in research work concerning concurrent and distributed programming
languages.

One of the challenges in the benchmark, henceforth called \emph{linearity
challenge}, concerns the formalisation of a \emph{minimal calculus of sessions}.
%
Sessions and session types~\citep{Honda93,HondaVasconcelosKubo98,HuttelEtAl16}
are established abstractions for the static analysis of distributed programs
based on peer-to-peer communication channels. Every session type system revolves
around three key ideas: (1) session endpoints are \emph{linear resources} that
cannot be discarded or duplicated without compromising some safety and liveness
properties of a program; (2) the type of a session endpoint is \emph{updated}
after each use to reflect the state of the protocol it describes; (3) peer
session endpoints are meant to be used in complementary ways so as to guarantee
the absence of communication errors and, to some extent, progress of the
interaction.
%
The linearity challenge is motivated by the observation that the proper
management of linear resources in a formalisation often requires a large number
of auxiliary definitions and technical results that divert from the main problem
under study~\cite{CarboneEtAl24}. One of the alleged culprits of such complexity
is \emph{context splitting}, namely the operation that partitions a typing
context in such a way that the linear resources described therein end up in only
one of the partitions. This observation has led to the exploration of various
alternative techniques including leftover typing~\cite{ZalakainDardha21}, the
use of linearity predicates~\cite{SanoKavanaghPientka23} and tagged
contexts~\cite{ZackonEtAl25}.

In this paper we approach the linearity challenge from a different angle:
\emph{instead of proposing new techniques that make it easy to formalise the
calculus in the challenge, we propose a (relatively) new calculus that is easy
to formalise with the existing techniques}.
%
More specifically, we describe a calculus called \LCP whose type system
coincides with the proof system of classical linear logic and that features
\emph{linear channels} instead of sessions. Linear channels are meant to be used
for a \emph{single} communication. Structured communications involving the
exchange of several messages within the same session are modeled in \LCP using
\emph{explicit continuations} according to the well-known encoding of (binary)
sessions into the linear
$\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
%
Overall, \LCP is a linear version of \CP~\cite{Wadler14} in which sessions are
encoded instead of being featured natively.
